The Spatial and Temporal Domains of Modern Ecology

### 2017-12-18 paper/main.tex
Editing main text through spatio-temporal domains. Updarted figure 1 vignette to calculate observational type percents by bins with unreplicated values removed.

\noindent \textbf{Results}\\
\noindent \textbf{\emph{Observational methods}}\\
To account for potential differences in scales related to methodology, we classified each observation according to the following broad categories, which were field methods (manual \emph{in situ} data collection), automated (\emph{in situ}) sensing, remote sensing/other geographic data, and paleo-reconstruction approaches. Field methods were used for 80\% of observations, automated sensing for 12.4\%, remote sensing for 6.9\%, and paleo-reconstruction for less than 0.8\%.

\noindent \textbf{\emph{Domains}}\\
In terms of resolution, the majority (67\%) of observations were collected in plots of $<$1 m$^2$ resolution, 24\% were collected within plots of 1 m$^2$ up to 1 ha, and the remaining 9\% in plots of $\geq$1 ha (Fig. 1A). The extent of 19\% of observations was $<$10 ha, 23\% covered 10-1,000 ha, 12\% 1,000-10,000 ha, 19\% 10,000-100,000 ha, 12\% 100,000-1,000,000 ha, and 15\% $>$1,000,000 ha (Fig. 1B).

@@

In the temporal dimensions, 37\% of observations were not repeated (Fig. 1C), 17\% were repeated at short intervals (sub-second to daily), 20\% at daily to monthly intervals, 18\% at monthly to yearly intervals, 6\% at yearly to decadal intervals, and 2\% at decadal or greater intervals. Duration was one day or less for 31\% of sampled observations (due to lack of temporal replication), while 10\% covered one day to one month, 23\% lasted one month to one year, 27\% covered 1-10 years, and 9\% spanned a decade or more (including several paleoecological studies covering centuries to millennia; Fig. 1D).

Juxtaposing these observational dimensions provides further insight into the spatio-temporal domains of observations (Fig. 2). Contrasting resolution with interval reveals that the majority of temporally replicated observations (unrepeated observations were excluded because they lack interval values) had resolutions of 10 cm$^2$-1 m$^2$ and were revisited at daily to yearly intervals (Fig. 2A). A less dense, oblong concentration of observations bounded on the lower right by monthly to yearly observations at 100 m$^2$ and on the upper left by near-daily to monthly observations with 1-10 ha resolution is also evident. Different observational methods occupied substantially different portions of the domain space, as indicated by the locations of the median resolution and interval values for each:  field observations had a resolution of 0.1-1 m$^2$ and were repeated at monthly intervals, whereas remote observations had coarser resolutions (1000 m$^2$) but finer repeat intervals ($\sim$1 day). Paleo-reconstructions and automated sensing techniques were both finely resolved (10 cm$^2$ to 0.01 m$^2$), but automated approaches had hourly-daily intervals while paleo-reconstructions had multi-decadal intervals.  The complete domain space of each observational type is detailed in the SI.

@@

Contrasting duration and extent (for all observations) reveals two primary domains of observational concentration. The first consists of observations spanning one month to one decade in time and 10-1000 ha in space, while the second is defined by observations of one year to several decades that cover 10,000 to 1,000,000 ha (Fig. 2B).  Three other notable, but lesser areas of concentration are also evident, including small area observations (0.1-1 ha) covering one month to decade, and short duration, temporally unreplicated observations ($<$1 day) of either 1-10 ha or 10,000-1,000,000 ha.

Comparing the two spatial dimensions against one another (for all observations) shows a primary concentration of observations with 10 cm$^2$ to 100 m$^2$ resolution that have extents ranging between slightly over 1,000 to nearly 1,000,000 ha (Fig. 2C). The second-most prominent concentration consists of higher resolution (1 cm$^2$-1 m$^2$), smaller extent (10-1,000 ha) observations, beneath which lies a third and fainter concentration of 1-1,000 cm$^2$ resolution, 1000 m$^2$ to $<$10 ha. These three concentrations suggest a tendency for observational extent to increase with resolution, which is a relationship that becomes more pronounced in the less densely observed portion of the domain where resolutions $\geq$100 m$^2$.

Most temporally replicated observations have daily to decadal intervals and durations of $\geq$ 1 month to 1 decade. The orientation of this concentration shows that interval increases with duration; observations lasting one month to one year tend to have daily to monthly intervals, while those lasting one year to one decade tend to have yearly to decadal intervals. The low densities of observations having sub-daily intervals shows that relatively few high frequency, long duration ecological measurements are undertaken.

For our final analysis of observational domains, we evaluated the differences between the scales represented by the extent and duration dimensions and the scales that ecological observations actually cover. To make this assessment, we log$_{10}$ transformed and then subtracted the values of i) actual extent from extent and ii) actual duration from duration, yielding the magnitude of difference between each pair of dimensions for each observation. We plotted these values (summarized in box plots) against their corresponding actual extent/duration values to evaluate whether these differences vary with scale (Fig. 3). This analysis showed that most (81\%) of the assessed observations had actual extents of $\leq$1 ha, which on average were 4 to nearly 8 orders of magnitude smaller than the primary measure of extent (Fig 3A). These differences declined linearily with actual extent, becoming approximately equal for the 5\% of observations having $\geq$100,000 ha actual extent. The actual duration of 64\% of observations was $<$1 day, which on average was 3 to nearly 9 orders of magnitude shorter than the time span covered by the conventional metric of duration (Fig. 3B). As with extent, the average magnitude of difference declined steadily with actual duration, becoming effectively equal for the 16\% of observations with $\geq$1 month of actual duration.

>>>>

\noindent \textbf{Results}\\
\noindent \textbf{\emph{Observational methods}}\\
To account for potential differences in scales related to methodology, we classified each observation according to the following broad categories, which were field methods (manual \emph{in situ} data collection), automated (\emph{in situ}) sensing, remote sensing/other geographic data (hereafter remote observations), and paleo-reconstruction approaches. Field methods were used for 80\% of observations, automated sensing for 12.4\%, remote sensing for 6.9\%, and paleo-reconstruction for less than 0.8\%.

\noindent \textbf{\emph{Distributions within the four primary dimensions}}\\
In terms of resolution, the majority (67\%) of observations (across all methods) were collected in plots of $<$1 m$^2$ resolution, 24\% were collected within plots of 1 m$^2$ up to 1 ha, and the remaining 9\% in plots of $\geq$1 ha (Fig. 1A). These distributions primarily reflect those of field observations, the dominant observational methodology. Examining the distributions for each observational method (Fig. S1 in SI) shows that automated sensing and paleo-reconstruction had resolutions that were generally finer (85\% or more $<$0.1 m$^2$) than field observations (47\% $<$0.1 m$^2$), while the majority of remote observations were much coarser (70\% $>$100 m$^2$).

The extent of 19\% of observations was $<$10 ha, 23\% covered 10-1,000 ha, 12\% 1,000-10,000 ha, 19\% 10,000-100,000 ha, 12\% 100,000-1,000,000 ha, and 15\% $>$1,000,000 ha (Fig. 1B).  As with resolution, the extent covered by automated sensing methods tended to be smaller (52\% $<$100 ha) than those of field observations (31\% $<$100 ha), while all but 4\% of remote observations covered areas $\geq$10,000 ha (as did the small number of paleo-reconstructions).

@@

In the temporal dimensions, 37\% of observations were not repeated (Fig. 1C), 17\% were repeated at short intervals (sub-second to daily), 20\% at daily to monthly intervals, 18\% at monthly to yearly intervals, 6\% at yearly to decadal intervals, and 2\% at decadal or greater intervals. With respect to temporally replicated observation (Fig. S1 in SI), automated sensing techniques had the finest intervals (61\% $\leq$1 day; 100\% $\leq$1 year), followed by remote observation (37\% $\leq$1 day; 78\% $\leq$1 year), field observations (17\% $\leq$1 day; 86\% $\leq$1 year), and paleo-reconstructions (21\% $\leq$1 decade).

Duration was one day or less for 31\% of sampled observations (due to lack of temporal replication), while 10\% covered one day to one month, 23\% lasted one month to one year, 27\% covered 1-10 years, and 9\% spanned a decade or more (including several paleoecological studies covering centuries to millennia; Fig. 1D). Paleo-reconstructions naturally had the longest duration (67\% $>$ 1 decade), while just $\sim$40\% of field, automated, and remote observations had durations of 1 year or longer.

\noindent \textbf{\emph{Spatial and temporal domains}}\\
Juxtaposing these observational dimensions provides further insight into the spatial and temporal domains of observations (Fig. 2). Contrasting resolution with interval reveals that the majority of temporally replicated observations (unrepeated observations were excluded because they lack interval values) had resolutions of 10 cm$^2$-1 m$^2$ and were revisited at daily to yearly intervals (Fig. 2A). A less dense, oblong concentration of observations bounded on the upper left by monthly to yearly observations at 100 m$^2$ resolution and on the lower right by near-daily to monthly observations with 1-10 ha resolution is also evident. The four observational methods occupied substantially different portions of the domain space, as indicated by the locations of their median values (and illustrated further in Fig. S2 in the SI): the median domain of field observations was between 0.1-1 m$^2$ of resolution with a monthly interval, whereas remote observations had coarser median resolutions (1000 m$^2$) but finer median intervals ($\sim$1 day). Paleo-reconstructions and automated sensing techniques were both finely resolved (medians from 10 cm$^2$ to 0.01 m$^2$), but automated approaches had hourly-daily median intervals compared to multi-decadal for paleo-reconstructions.

@@

Comparing the interval and duration of temporally replicated observations showed most observations had daily to decadal intervals and durations of $\geq$1 month up to 1 decade (Fig. 2B). The orientation of this concentration shows that interval increases with duration; observations lasting one month to one year tend to have daily to monthly intervals, while those lasting one year to one decade tend to have yearly to decadal intervals. This tendency is reflected in the median domain locations of the primary observational methods: automated sensing had the finest median interval (hour-day) and shortest duration (month-year), followed by remote sensing (median interval slightly greater than one day and median duration of 1 year), field observations (median monthly interval and duration just over 1 year), and finally paleo-reconstructions (median interval 1 decade and millennial duration). The low densities of observations having sub-daily intervals shows that relatively few high frequency, long duration ecological measurements are undertaken; this position

Contrasting the two spatial dimensions against one another (for all observations) shows a primary concentration of observations with 10 cm$^2$ to 100 m$^2$ resolution with extents ranging from just over 1,000 to nearly 1,000,000 ha (Fig. 2C). The second-most prominent concentration consists of higher resolution (1 cm$^2$-1 m$^2$), smaller extent (10-1,000 ha) observations, beneath which lies a third and fainter concentration of 1-1,000 cm$^2$ resolution, 1000 m$^2$ to $<$10 ha. These three concentrations suggest a tendency for observational extent to increase with resolution, which is further evident in the median domain values (and kernel densities; Fig. S2 in SI) of automated (0.01 m$^$ resolution, 100 ha extent), field (0.1-1 m$^2$ resolution, 1,000-10,000 ha extent), and remote (1,000 m$^2$ resolution, 1-10 million ha extent) observations.  Paleo-reconstructions were an outlier from this relationship, having very fine median resolution (0.01 m$^2$) but large extent (1 million ha), a result that likely reflects the very low sample size for this observational type.

Two primary domains of observational concentration are revealed by juxtaposing duration and extent (across all observations). The first consists of observations spanning one month to one decade in time and 10-1,000 ha in space, while the second is defined by observations of one year to several decades that cover 10,000 to 1,000,000 ha (Fig. 2D).  Three other notable, but lesser areas of concentration are also evident, including small area observations (0.1-1 ha) covering one month to decade, and short duration, temporally unreplicated observations ($<$1 day) of either 1-10 ha or 10,000-1,000,000 ha.  The median observation (1 year duration, 100 ha extent) from automated sensing lies near the center of the second major concentration, while the median extents of remote (1-10 million ha) and field observations bound the primary concentration at its upper and lower extents, while the median duration of both observational types falls between 1 month to 1 year.

\noindent \textbf{\emph{Differences between actual and ostensible scales}}\\
For our final analysis of observational domains, we evaluated the differences between the scales represented by the extent and duration dimensions and the scales that ecological observations actually cover. To make this assessment, we log$_{10}$ transformed and then subtracted the values of i) actual extent from extent and ii) actual duration from duration, yielding the magnitude of difference between each pair of dimensions for each observation. We plotted these values (summarized in box plots) against their corresponding extent/duration values to evaluate whether these differences vary with scale (Fig. 3). This analysis showed that most (81\%) of the assessed observations had actual extents of $\leq$1 ha, which on average were 4 to nearly 8 orders of magnitude smaller than the primary measure of extent (Fig 3A). These differences declined linearily with actual extent, becoming approximately equal for the 5\% of observations having $\geq$100,000 ha actual extent. The actual duration of 64\% of observations was $<$1 day, which on average was 3 to nearly 9 orders of magnitude shorter than the time span covered by the conventional metric of duration (Fig. 3B). As with extent, the average magnitude of difference declined steadily with actual duration, becoming effectively equal for the 16\% of observations with $\geq$1 month of actual duration.
